Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-07-20 ebayes_rnaseq_heterosis.tex
more Dan edits
blinded version

\newcommand{\blind}{0} % 0 = not blind, 1 = blind

@@

\date{{\it Received December} 2014}

@@

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. This article has supplementary material online.
\end{abstract}

@@

We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
\psi_g \stackrel{ind}{\sim} N(0,\sigma_\psi^2). 
\]
To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.

@@

Conditional on the estimated normalization factors $\hat{\gamma}$ and hyperparameters $\hat{\pi}$, we perform a Bayesian analysis to re-estimate the gene-specific parameters and describe their uncertainty. Conditional on $\hat{\gamma}$ and $\hat{\pi}$ the gene-specific parameters are independent, i.e.
\begin{equation}
\begin{split}
p\left(\theta\left|y,\hat{\pi},\hat{\gamma}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
&\times 
\left. \vphantom{\prod_{v=1}^V}
N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
\right]
\label{e:condind}
\end{split}
\end{equation}
Thus, conditional posterior inference across the genes can be obtained independently and in parallel.

@@

\ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}. One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density. Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid difference ($\delta_g$) and instead use normal priors in this situation. An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior. Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}. Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for $\hat{\delta}_g$ and $\hat{\alpha}_g$, e.g. 
\[ 
% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
% 0 & \mbox{otherwise}
% \end{array} \right.
P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \ge 0\right).
%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
%0 & \mbox{otherwise}
%\end{array} \right.
\]
HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

\section*{Supplementary Materials}

The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).  
%the all-genes-independent (\verb@model_ind.txt@) and covariance (\verb@model_cov.txt@) models used in obtaining MAP estimates, along with the single-gene-independent (\verb@sg_model_ind.txt@) and covariance (\verb@sg_model_cov.txt@) models used in performing the MCMC on single genes.

>>>>

\newcommand{\blind}{1} % 0 = not blind, 1 = blind

@@

\date{{\it Received July} 2015}

@@

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (\RNAseq{}) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for \RNAseq{} analysis. To address these shortcomings, we develop a hierarchical negative binomial model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. 
% This article has supplementary material online.
\end{abstract}

@@

We assume a hierarchical model for the gene-specific mean parameters and overdispersion parameters. The parental averages and overdispersion parameters are assumed to follow a normal distribution, i.e.
\[ 
%\gamma_{vi} \stackrel{ind}{\sim} N(0,\sigma_\gamma^2), \quad 
\phi_g \stackrel{ind}{\sim} N(\eta_\phi, \sigma_\phi^2), \quad \mbox{and}\quad 
\psi_g \stackrel{ind}{\sim} N(\eta_\psi,\sigma_\psi^2). 
\]
To allow for more shrinkage with the half-parental and hybrid differences, we use a Laplace (or double exponential prior) and thus implement a Bayesian LASSO \citep{park2008bayesian}, i.e. 
\[ 
\alpha_g \stackrel{ind}{\sim} La(\eta_\alpha, \sigma_\alpha) \qquad \mbox{and} \qquad
\delta_g \stackrel{ind}{\sim} La(\eta_\delta, \sigma_\delta) 
\]
where $\alpha\sim La(\eta,\sigma)$ has a probability density function given by $La(\alpha;\eta,\sigma) = \exp(-|\alpha-\eta|/\sigma)/2\sigma$ with expectation $\eta$ and variance $2\sigma^2$. We assume independence amongst the parental averages, half parental differences, hybrid differences, and overdispersion parameters.

@@

Conditional on the estimated normalization factors $\hat{\gamma}$ and hyperparameters $\hat{\pi}$, we perform a Bayesian analysis to re-estimate the gene-specific parameters and describe their uncertainty. Conditional on $\hat{\gamma}$ and $\hat{\pi}$ the gene-specific parameters are independent, i.e.
\begin{equation}
\begin{split}
p\left(\theta\left|y,\hat{\pi},\hat{\gamma}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
%&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right.\right) \right] p(\theta_g|\hat{\eta},\hat{\sigma}) \\
\propto \prod_{g=1}^G & \left[ \,\prod_{v=1}^V \prod_{i=1}^{n_v} \mbox{NB}\left(y_{gvi};e^{\mu_{gv}+\hat{\gamma}_{vi}},e^{\psi_g}\right)  \right. \\
&\times 
\left. \vphantom{\prod_{v=1}^V}
N(\phi_g;\hat{\eta}_\phi, \hat{\sigma}_\phi^2)
La(\alpha_g;\hat{\eta}_\alpha, \hat{\sigma}_\alpha)
La(\delta_g;\hat{\eta}_\delta, \hat{\sigma}_\delta)
N(\psi_g;\hat{\eta}_\psi, \hat{\sigma}_\psi^2) 
\right]
\label{e:condind}
\end{split}
\end{equation}
Thus, conditional posterior inference across the genes can be obtained independently and in parallel.

@@

In addition to the approach above, we utilized \ShrinkBayes{} to estimate heterosis probabilities with two modifications described here. \ShrinkBayes{} utilizes integrated nested Laplace approximation (INLA) \citep{rue2009approximate} in combination with empirical Bayes ideas \citep{van2014shrinkbayes}. One limitation with inferential methods based on INLA is that all distributions, except for the data distribution, must have tails as light or lighter than the normal density. Thus, we cannot implement the Laplace priors for the half-parental difference ($\alpha_g$) and the hybrid difference ($\delta_g$) and instead use normal priors in this situation. An additional limitation is that INLA provides approximations to marginal posteriors for parameters or linear combinations of parameters, but not an approximation to the full posterior. Since we are interested in non-linear quantities such as $P(\delta_g > |\alpha_g||y)$, we cannot compute these directly using \ShrinkBayes{}. Instead, for \ShrinkBayes{}, we calculate heterosis probabilities conditional on posterior means for $\hat{\delta}_g$ and $\hat{\alpha}_g$, e.g. 
\[ 
% P_{\ShrinkBayes{}}(LPH_g|y) \equiv \left\{ \begin{array}{ll}
% P(\delta_g>-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
% P(\delta_g>\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
% 0 & \mbox{otherwise}
% \end{array} \right.
P_{\ShrinkBayes{}}(\mbox{LPH}_g|y) \equiv P(\delta_g<-\alpha_g|y)\mathrm{I}\left(\hat{\delta}_g < -\hat{\alpha}_g \le 0\right) + P(\delta_g<\alpha_g|y) \mathrm{I}\left(\hat{\delta}_g < \hat{\alpha}_g \ge 0\right).
%P_{\ShrinkBayes{}}(HPH|y) \equiv \left\{ \begin{array}{ll}
%P(\delta_g<-\alpha_g|y) & \mbox{if } \hat{\delta}_g > -\hat{\alpha}_g \ge 0 \\
%(\delta_g<\phantom{-}\alpha_g|y) & \mbox{if } \hat{\delta}_g > \phantom{-}\hat{\alpha}_g \ge 0 \\
%0 & \mbox{otherwise}
%\end{array} \right.
\]
HPH probability is defined similarly with all inequalities reversed. As before, we construct a ranked list of genes according to the maximum of the gene's LPH and HPH heterosis probabilities.
%the heterosis probability for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g>\alpha_g|y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.

@@

%\section*{Supplementary Materials}

%The supplementary materials include three plain text files: the maize experiment data ({\tt data.csv}), a script to run our method ({\tt script.R}) on this data, and the \Stan{} model ({\tt model.stan}).
